---
title: "My United States Travel Map"
author: Daniel Jang
date: "`r format(Sys.time(), '%B %d, %Y')`"
date-format: iso
format: 
  html: 
    embed-resources: true
execute:
  echo: false
  warning: false
  error: false
  message: false
editor: source
---

```{r}
library(tidyverse)
library(tidyselect)
library(tibble)
library(lubridate)
library(sf)
library(tigris)
library(lubridate)
library(tidygeocoder)

library(gt)
library(gtsummary) # nice count visualizations
library(forcats) # a specific data wrangling issue
library(readxl)

library(wordcloud)
library(rstatix)
```

This is still a work in progress. As I add new visualizations, some other things may be broken, so please excuse the state of this html file. This also serves as an exercise in data visualization.

```{r}
# Regions defined by the U.S. census bureau
northeast <- c("Massachusetts", "New Hampshire", "Vermont", "Connecticut", "Maine", "Rhode Island", 
               "New York", "New Jersey", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", 
             "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota")

south <- c("Delaware", "DC", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia",
           "Alabama", "Kentucky", "Mississippi", "Tennessee",
           "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", 
          "Alaska", "California", "Hawaii", "Oregon", "Washington")
```

```{r, output = FALSE}
# data that I type in manually
daniel_travel_and_lived_map <- readxl::read_xlsx("trip_and_lived_data.xlsx") |> # I can also add coordinate information here... maybe play around with that?
  mutate(start_date = as_date(start_date),
         end_date   = as_date(end_date)) |>
  mutate(region = case_when(
    state %in% northeast ~ "Northeast",
    state %in% midwest ~ "Midwest",
    state %in% south ~ "South",
    state %in% west ~ "West"
    )) |>
  mutate(travel_length = as.integer(end_date - start_date) + 1) |>
  mutate(travel_year = year(start_date))

# separate dataset without lived data
daniel_travel_map <- daniel_travel_and_lived_map |> 
  filter(type != "Lived") |>
  drop_na() #any NA is further removed

# these are places I don't want included
us_territories <- c("VI", "GU", "MP", "AS")

# US states shapefile
us <- states(cb = TRUE, resolution = "5m") |>
  filter(!(STUSPS %in% us_territories)) |>
  shift_geometry() #for alaska and hawaii (and PR)
```

## Which states have I been to in my lifetime (both lived and travelled)?

My definition of "lived in a state" is if I spent at least 31 consecutive days in the same semi-permanent lodging, enough to develop a routine. For travelling, I do not count passing by a state in some form of transportation, but I do count some destination, even if it was unintentional. For example, I originally had a layover in Phoenix, Arizona, but I ended up staying the night. 

> As a bonus, I will denote states whose capital city I have visited with a solid line. 

```{r}
# Counts
capital_counts <- daniel_travel_and_lived_map |>
  filter(capital_indicator == "Yes") |>
  distinct(state) |> as_vector()

lived_counts <- daniel_travel_and_lived_map |>
  filter(type == "Lived") |>
  distinct(state) |> as_vector()

visit_counts <- daniel_travel_and_lived_map |>
  count(state, name = "visits") |>
  mutate(capital_indicator = ifelse(state %in% capital_counts, "Yes", "No"),
         lived_indicator = ifelse(state %in% lived_counts, "Yes", "No"))

# Also gather information from "lived" dataset
```

```{r}
# Join counts to map
travel_count_map <- us |>
  left_join(
    visit_counts,
    by = c("NAME" = "state")
  ) |>
  mutate(
    visits = ifelse(is.na(visits), "No", visits), # I did not visit any state in which I have no data for
    lived = ifelse(is.na(lived_indicator), "No", lived_indicator), # same idea as above but for living
    visit_indicator = ifelse(visits == "No", "No", "Yes"),
    capital_indicator = ifelse(is.na(capital_indicator), "No", capital_indicator) 
  ) |>
  mutate(
    visit_and_lived_indicator = ifelse(
      !is.na(lived_indicator) & lived_indicator == "Yes", #to handle NA case
      "Lived", 
      visit_indicator) #more complete version)
    ) 
```

```{r, fig.width=18, fig.height=12}
# Plot
ggplot(data = travel_count_map) +
  geom_sf(aes(fill = factor(visit_and_lived_indicator), linetype  = factor(capital_indicator)),
          linewidth = 1) +
  scale_fill_manual(
    values = c("No" = "white", "Yes" = "lightblue", "Lived" = "cyan"), 
    name = "Visited State") +
  scale_linetype_manual(
    values = c("No" = "dotted", "Yes" = "solid"),
    name = "Visited Capital"
  ) +
  theme_bw() +
  theme(
    axis.text = element_blank(),
    panel.grid = element_blank(),
    legend.position = "bottom") +
  labs(
    title = "Have I been to the state?"
  )
```

## Step Visualization of first visit (since 2018)

Of course I did not visit these states at the same time, so I wanted to see when I visited each state for the first time. From this point on, I only use data after 2018 since I do not have accurate data prior to this time.

```{r}
# first, count unique states from the earliest I've been there
first_travel <- daniel_travel_map |>
  group_by(state) |>
  filter(start_date == min(start_date)) |> # somehow this worked lol
  rename(first_visit_date = start_date) |> 
  ungroup(state)

first_travel <- first_travel |>
  mutate(states_visited_order = nrow(first_travel) - row_number() + 1) #again, stupid fix but it works

# now we can plot it
ggplot(data = first_travel, aes(x = first_visit_date, y = states_visited_order)) +
  geom_step() + 
  geom_text(aes(label = state, color = region), nudge_x = -225, size = 3) + 
  theme_bw() +
  labs(x = "Date of First Travel (incomplete)",
       y = "Number of States",
       title = "States visited for the first time",
       color = "Region")
```

## Visits (days) and living each year (since 2018)

As an exercise in small multiple visualizations, I have here a map from each year highlighting how long I stayed in each state I visited each year (by number of days). I also added a border highlight for any state that I lived in during that time. 

```{r}
# first, loop through the minimum and maximum years of my dataset
year_min <- min(daniel_travel_and_lived_map$travel_year)
year_max <- max(daniel_travel_and_lived_map$travel_year)

# I'm going to vectorize this in a minute
travel_duration_map <- purrr::map_dfr(year_min:year_max, \(yr) {
  daniel_travel_and_lived_map |>
    filter(year(start_date) <= yr & year(end_date) >= yr) |> 
    right_join(us, by = c("state" = "NAME")) |>
    mutate(
      travel_year = yr,
      travel_length = coalesce(travel_length, 0),
      lived_indicator = case_when(type == "Lived" ~ "Yes",
                                  TRUE ~ "No")
    )
})

# some quick selection to make data process cleaner
travel_duration_map <- travel_duration_map |>
  filter(travel_year >= 2018) |>
  mutate(travel_length = case_when(
    lived_indicator == "Yes" ~ 0, #this is for lived data only
    TRUE ~ travel_length)) |>
  dplyr::select(state, travel_length, travel_year, lived_indicator, STUSPS, geometry) |> 
  st_as_sf() 
```


```{r, fig.width=18, fig.height=12}
# the plot!
ggplot(data = travel_duration_map) +
  geom_sf(aes(fill = travel_length, color = factor(lived_indicator))) +
  scale_fill_viridis_c(name = "Trips") +
  scale_fill_gradient(low = "white", high = "blue") + 
  scale_color_manual(
    values = c("No" = "grey40", "Yes" = "orange"),
    labels = c("No", "Yes")
  ) + 
  facet_wrap(~ travel_year) + 
  theme_bw() + theme(axis.text = element_blank(), 
                     panel.grid = element_blank(),
                     legend.position = "bottom") +
  labs(title = "Duration of stay in each state per year (since 2018)",
       fill = "Length of Travel (Days)",
       color = "Lived here?")
```

## When did I visit each state? 

This is kind of obselete and not that interesting in my opinion, so I may get rid of it.

```{r}
state_dates <- data.frame(
  state = c(
    "Delaware","Pennsylvania","New Jersey","Georgia","Connecticut","Massachusetts",
    "Maryland","South Carolina","New Hampshire","Virginia","New York","North Carolina",
    "Rhode Island","Vermont","Kentucky","Tennessee","Ohio","Louisiana","Indiana",
    "Mississippi","Illinois","Alabama","Maine","Missouri","Arkansas","Michigan",
    "Florida","Texas","Iowa","Wisconsin","California","Minnesota","Oregon",
    "Kansas","West Virginia","Nevada","Nebraska","Colorado","North Dakota",
    "South Dakota","Montana","Washington","Idaho","Wyoming","Utah","Oklahoma",
    "New Mexico","Arizona","Alaska","Hawaii","DC"
  ),
  date_joined = as.Date(c(
    "1787-12-07","1787-12-12","1787-12-18","1788-01-02","1788-01-09","1788-02-06",
    "1788-04-28","1788-05-23","1788-06-21","1788-06-25","1788-07-26","1789-11-21",
    "1790-05-29","1791-03-04","1792-06-01","1796-06-01","1803-03-01","1812-04-30",
    "1816-12-11","1817-12-10","1818-12-03","1819-12-14","1820-03-15","1821-08-10",
    "1836-06-15","1837-01-26","1845-03-03","1845-12-29","1846-12-28","1848-05-29",
    "1850-09-09","1858-05-11","1859-02-14","1861-01-29","1863-06-20","1864-10-31",
    "1867-03-01","1876-08-01","1889-11-02","1889-11-02","1889-11-08","1889-11-11",
    "1890-07-03","1890-07-10","1896-01-04","1907-11-16","1912-01-06","1912-02-14",
    "1959-01-03","1959-08-21","1790-07-16"  # DC last
  ))
)

daniel_travel_map_ordered <- daniel_travel_map |>
  mutate(state = factor(state, levels = state_dates$state[order(state_dates$date_joined, decreasing = TRUE)]))

ggplot(data = daniel_travel_map_ordered, aes(x = start_date, y = state, color = region)) +
  geom_point() + theme_bw() + 
  labs(x = "Year",
       y = "State", 
       color = "Region",
       title = "Every Visit ordered by when the state joined the union")
``` 

## Wordcloud of Travel Descriptions!

Turns out I am not that creative in my writing. 

```{r}
library(tidytext)
library(stopwords)

my_stopwords <- stopwords("en", source = "stopwords-iso")

travel_wordcloud <- daniel_travel_map |> # don't need lived data
  tidytext::unnest_tokens(word, desc) |> # breaks text columns into single words
  count(word, sort = TRUE) |>
  rename(freq = n) |> # rename for clarity
  filter(!(word %in% my_stopwords), #remove any stopwords
         freq > 1) # word should appear more than once 

# the wordcloud!
wordcloud(words     = travel_wordcloud$word, 
          freq      = travel_wordcloud$freq, 
          min.freq  = 150, 
          max.words = 150, 
          random.order=FALSE, rot.per=0.3, colors=brewer.pal(8, "Dark2"))
```

## Descriptive Statistics

Things that I am planning to include: 

1. how many redeyes?
2. how many business vs leisure trips?
3. distribution of days I stayed (faceted by something)?
4. average length of trip by days?
5. number of day trips?
6. average distance of trips?
7. I also may change the fourth visualization into a chart instead, with states ordered by when they joined the union.
8. Also something with cities (geographic coordinates? I wonder if there's a public data set to do this with... I'm confident there is)

```{r, results = FALSE}
ggplot(data = daniel_travel_map, aes(x = travel_length)) +
  geom_histogram() + theme_bw()
```

## The following are visuals I am unsure if I should make.

1. Lines of travel (How have I travelled between each state?): May not work if I don't have proper coordinates. Maybe I can do it based on thickness? *I am unconvinced of this one* `If I can combine this with city data, then that would be really cool, but that sounds like so much effort`.

2. A county map? Includes places I've just passed by: something approximate, since the map version here is kind of crazy

> I may just do something from mapchart.net

```{r, output = FALSE}
us_counties <- counties(cb = TRUE, class = "sf") |> 
  filter(!(STUSPS %in% us_territories)) |>
  shift_geometry()
```

```{r}
ggplot(data = us_counties) +
  geom_sf(fill = "white", color = "gray60", linewidth = 0.1) +
  theme_void() +
  labs(title = "All US Counties")
```
