---
title: "My Travel Map"
author: Daniel Jang
date: "`r format(Sys.time(), '%B %d, %Y')`"
date-format: iso
format: 
  html: 
    embed-resources: true
execute:
  echo: false
  warning: false
  error: false
  message: false
editor: source
---

```{r}
library(tidyverse)
library(tidyselect)
library(tibble)
library(lubridate)
library(sf)
library(tigris)
library(lubridate)
library(tidygeocoder)

library(gt)
library(gtsummary) # nice count visualizations
library(forcats) # a specific data wrangling issue
library(readxl)
```

THIS IS STILL A MAJOR WORK IN PROGRESS. I JUST HAVE THE RAW FORMAT HERE.
> I need to add Canadian trips to my dataset and exclude them in the United States portion (or maybe incorporate? I don't know at the moment)

# The United States

```{r}
# Regions defined by the U.S. census bureau
northeast <- c("Massachusetts", "New Hampshire", "Vermont", "Connecticut", "Maine", "Rhode Island", 
               "New York", "New Jersey", "Pennsylvania")

midwest <- c("Illinois", "Indiana", "Michigan", "Ohio", "Wisconsin", 
             "Iowa", "Kansas", "Minnesota", "Missouri", "Nebraska", "North Dakota", "South Dakota")

south <- c("Delaware", "DC", "Florida", "Georgia", "Maryland", "North Carolina", "South Carolina", "Virginia", "West Virginia",
           "Alabama", "Kentucky", "Mississippi", "Tennessee",
           "Arkansas", "Louisiana", "Oklahoma", "Texas")

west <- c("Arizona", "Colorado", "Idaho", "Montana", "Nevada", "New Mexico", "Utah", "Wyoming", 
          "Alaska", "California", "Hawaii", "Oregon", "Washington")
```


```{r, output = FALSE}
# data that I type in manually
daniel_travel_map <- readxl::read_xlsx("trip_data.xlsx") |> # I can also add coordinate information here... maybe play around with that?
  mutate(start_date = as_date(start_date),
         end_date   = as_date(end_date)) |>
  mutate(region = case_when(
    state %in% northeast ~ "Northeast",
    state %in% midwest ~ "Midwest",
    state %in% south ~ "South",
    state %in% west ~ "West"
    )) |>
  mutate(travel_length = as.integer(end_date - start_date) + 1) |>
  drop_na() #any NA is removed

# US states shapefile
us <- states(cb = TRUE) |>
  shift_geometry() #for alaska and hawaii (and PR)
```

These visualizations do not take into account smaller trips within the state I live in (e.g. travelling from Allston to Boston doesn't count, but something like Haverhill to Fall River counts). These are just ones I thought were memorable. Thus, I don't count just passing through a state. I also don't count states I have lived in; that will probably be a separate visualization.

```{r}
# Count
visit_counts <- daniel_travel_map |>
  count(state, name = "visits")

# Join counts to map
map_data <- us |>
  left_join(
    visit_counts,
    by = c("NAME" = "state")
  ) |>
  mutate(
    visits = ifelse(is.na(visits), 0, visits)
  )

# Plot
ggplot(map_data) +
  geom_sf(aes(fill = visits)) +
  coord_sf(xlim = c(-3500000, 3500000), ylim = c(-2500000, 2500000), expand = FALSE) + #stupid fix by playing around with numbers
  scale_fill_viridis_c(name = "Trips") +
  scale_fill_gradient(low = "white", high = "blue") + 
  theme_minimal() +
  labs(title = "How many times have I made significant travel to this state?")
```

2. Timeline on unique states I have been to (manually add states I lived in like California), and on the X axis, show the state name.

> Still a couple of fixes I want here since it's not 100% accurate but it's fine

```{r}
# first, count unique states from the earliest I've been there
first_travel <- daniel_travel_map |>
  group_by(state) |>
  filter(start_date == min(start_date)) |> # somehow this worked lol
  rename(first_visit_date = start_date) |> 
  ungroup(state)

first_travel <- first_travel |>
  mutate(states_visited_order = nrow(first_travel) - row_number() + 1) #again, stupid fix but it works

# now we can plot it
ggplot(data = first_travel, aes(x = first_visit_date, y = states_visited_order)) +
  geom_step() + 
  geom_text(aes(label = state, color = region), nudge_x = -225, size = 3) + 
  theme_bw() +
  labs(x = "Date of Travel",
       y = "Number of States",
       title = "States visited for the first time",
       color = "Region")
```

3. Small multiples visualization of states I have visited each year + lived in (e.g. 2025 will have MA and NY as "lived" and RI, TX as travelled)
> I think this is more interesting in my head

```{r}

```

3.5. Approximately when did I visit each state?

```{r}
state_dates <- data.frame(
  state = c(
    "Delaware","Pennsylvania","New Jersey","Georgia","Connecticut","Massachusetts",
    "Maryland","South Carolina","New Hampshire","Virginia","New York","North Carolina",
    "Rhode Island","Vermont","Kentucky","Tennessee","Ohio","Louisiana","Indiana",
    "Mississippi","Illinois","Alabama","Maine","Missouri","Arkansas","Michigan",
    "Florida","Texas","Iowa","Wisconsin","California","Minnesota","Oregon",
    "Kansas","West Virginia","Nevada","Nebraska","Colorado","North Dakota",
    "South Dakota","Montana","Washington","Idaho","Wyoming","Utah","Oklahoma",
    "New Mexico","Arizona","Alaska","Hawaii","DC"
  ),
  date_joined = as.Date(c(
    "1787-12-07","1787-12-12","1787-12-18","1788-01-02","1788-01-09","1788-02-06",
    "1788-04-28","1788-05-23","1788-06-21","1788-06-25","1788-07-26","1789-11-21",
    "1790-05-29","1791-03-04","1792-06-01","1796-06-01","1803-03-01","1812-04-30",
    "1816-12-11","1817-12-10","1818-12-03","1819-12-14","1820-03-15","1821-08-10",
    "1836-06-15","1837-01-26","1845-03-03","1845-12-29","1846-12-28","1848-05-29",
    "1850-09-09","1858-05-11","1859-02-14","1861-01-29","1863-06-20","1864-10-31",
    "1867-03-01","1876-08-01","1889-11-02","1889-11-02","1889-11-08","1889-11-11",
    "1890-07-03","1890-07-10","1896-01-04","1907-11-16","1912-01-06","1912-02-14",
    "1959-01-03","1959-08-21","1790-07-16"  # DC last
  ))
)

daniel_travel_map_ordered <- daniel_travel_map |>
  mutate(state = factor(state, levels = state_dates$state[order(state_dates$date_joined, decreasing = TRUE)]))

ggplot(data = daniel_travel_map_ordered, aes(x = start_date, y = state, color = region)) +
  geom_point() + theme_bw() + 
  labs(x = "Year",
       y = "State", 
       color = "Region",
       title = "Every Visit ordered by when the state joined the union")
``` 

4. Word cloud of my travel descriptions

```{r, results = FALSE}
daniel_travel_map |> dplyr::select(desc)
```

5. Descriptive statistics: how many redeyes, how many business vs leisure trips, distribution of days I stayed (faceted by something), average length of trip by days, number of day trips, average distance of trips (MAYBE) etc.

```{r}
ggplot(data = daniel_travel_map, aes(x = travel_length)) +
  geom_histogram() + theme_bw()
```

6. Capital city list: Have I been to the state capital?

```{r}

```

7. Lines of travel (How have I travelled between each state?): May not work if I don't have proper coordinates. Maybe I can do it based on thickness?

8. A county map? Includes places I've just passed by (do this manually): something approximate.

